Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2013-10-30 01_intro.tex
Sections 2,3 proof

Despite the RSA model's theoretical promise and empirical success, if
we analyze the design of the model, we will find several choices that
are not \emph{prima facie} obvious and might thus need further
consideration and justification. These choices have their alternatives
in closely related game-theoretic approaches to pragmatics
(e.g. \cite{Benz2007,Jager2013}), which have similar but slightly
different conceptual motivations. Hence it is important to
systematically compare the original RSA model with all these
alternatives and test their predictions on empirical data, so as to
gain a better understanding of the relation between these
models. Doing so will help illuminate the theoretical commitments and
empirical implications of each design choice, which will enhance our
understanding of the nature of human pragmatic inference.

>>>>

Despite the RSA model's theoretical promise and empirical success, if
we analyze the design of the model, we will find several choices that
are not \emph{prima facie} obvious and might thus need further
consideration and justification. These choices have their alternatives
in closely related game-theoretic approaches to pragmatics
\cite{Rabin1990:Communication-b,Stalnaker:SayingMeaningCredibility,BenzvanRooijOptimalAssertions2007,Franke2011:Quantity-Implic,Jager2013:Rationalizable-}, which have similar but slightly
different conceptual motivations. Hence it is important to
systematically compare the original RSA model with all these
alternatives and test their predictions on empirical data, so as to
gain a better understanding of the relation between these
models. Doing so will help illuminate the theoretical commitments and
empirical implications of each design choice, which will enhance our
understanding of the nature of human pragmatic inference.

### 2013-10-30 02_RSA.tex
rephrased intro, amended layout
Sections 2,3 proof

Figure \ref{context} is an example of the referential game used in studying human pragmatic inference \cite{Frank}. In a referential game there is a set of objects called the \emph{context}. The speaker in the game is asked to refer to one of the objects (the \emph{target}) by uttering an expression (typically a word) to the listener. The listener, who has no idea which object is the target, needs to recover the speaker's intended referent based on the speaker's utterance.

\begin{figure}[h] 
  \centering
  \subfigure[The context]{\label{context}
  \begin{tikzpicture}
     \path (-1.8, 0) node [shape=rectangle, draw=green!65 ,fill= green!65 ,minimum size=28]{}
           ( 0, 0) node  [shape=circle, draw=green!65, fill= green!65,minimum size=30] {}
           ( 1.8, 0) node [shape=circle, draw=blue!65, fill=blue!65  ,minimum size=30] {}
           ( -1.8, -0.8) node {$t_{1}$}
           ( -0.0, -0.8) node {$t_{2}$}
           (  1.8, -0.8) node {$t_{3}$}
     ;

  \end{tikzpicture}
  }  
  \hspace{5mm}
  \subfigure[The vocabulary and the truth table]{\label{VaTT}
  \begin{tikzpicture}
     \path (0,0) node[] 
     {\begin{tabular}{r|ccc}
                         & \quad $t_{1}$\ \  &\quad $t_{2}$\ \  &\quad $t_{3}$ \ \   \\
    \hline
   ``Green'':  $m_{1}$ \quad  &  $1$        &   $1$       & $0$      \\
   ``Circle'': $m_{2}$  \quad &  $0$        &   $1$       & $1$      \\ 
   ``Square'': $m_3$ \quad    &  $1$      &   $0$       & $0$      \\
   ``Blue'':   $m_4$  \quad  &  $0$        &   $0$       & $1$     
     \end{tabular}
     } ;
  \end{tikzpicture}
  }
  \caption{A simple referential game}
\end{figure}

Let us first try to play this game ourselves, to get a better sense of what it is about and why it is relevant to human pragmatic reasoning. For simplicity let us assume that the utterances that the speaker is allowed to use are from a \emph{vocabulary} which is commonly known between the speaker and the listener. The content of the vocabulary and the (truth-conditional/literal) meaning of each word are shown in Figure \ref{VaTT}. Suppose we play as the speaker and the target is $t_1$, the green square. We can either use ``Square'' or ``Green'' to literally truthfully refer to it, but it seems more prudent to use ``Square'', as there is only one square in the context and thus the listener can easily identify it, whereas there are two green objects, which makes ``Green'' ambiguous. In terms of the Gricean \emph{Maxim of Quantity}, we should use ``Square'' because it is as maximally informative as is required. Similarly, we should use ``Blue'' to refer to $t_3$, the blue circle. However,  using ``Blue'' to refer to the blue circle might intuitively sound a little unnatural, as color terms are usually used as adjectives\footnote{They are used as nouns mostly to refer to the colors themselves.}, while we usually use nouns to refer to concrete objects. This inclination becomes more evident when we want to refer to $t_2$, the green circle. While ``Green'' and ``Circle'' are equally ambiguous in this case, we might nevertheless prefer the latter.

Now let us turn to play as the listener. If we hear ``Square'' or ``Blue'', we easily know the intended referent as there is no ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are two circles in the context that we need to choose from. On the one hand, the blue circle, having the unique color in the context, seems to be perceptually dominant and thus easily captures our attention. On the other hand, from the previous analysis we know that if the blue circle were the intended referent, the speaker could have chosen ``Blue'' which is not ambiguous and thus more informative. Hence the listener needs to balance two sources of information, i.e. the (presumably subconscious) perceptual salience of different objects and the rational expectation of the likelihood of the speaker making the utterance for each object. The latter line of reasoning is crucial to the classic Gricean account of scalar implicatures, and the major challenges in terms of formal modeling are how to quantify  notions such as informativeness and how different pieces of information should be integrated.

The RSA model addresses the problems by using information-theoretic concepts to measure informativeness and Bayesian inference to integrate different sources of information \cite{Frank}.

In order to measure the informativeness of an utterance, the RSA model starts with the literal listener who upon receiving an utterance $m$ does a conditioning on its literal meaning:
\begin{equation}\label{Bayesian_literal}
\rho_0(t \mid m) = \mathcal{U}(t\mid \intp{m}) ,
\end{equation}
where $\mathcal{U}$ is a uniform distribution over all the possible referents. Now the informativeness of utterance $m$ for the intended referent $t$ can be measured as the negative Kullback-Leibler divergence of the induced literal listener's belief $\rho_0$ from the speaker's own belief $\delta_t$:
\begin{equation} \label{Bayesian_sender_informativeness}
\mbox{Info}(m,t)=-\mbox{KL}(\delta_t \| \rho_0)=-\sum_{t'}\delta_t(t')\log\left(\frac{\delta_t(t')}{\rho_0(t')}\right),
\end{equation}
where $\delta_t$ is a delta distribution with all the probability mass on $t$, as the speaker knows her intended referent:
\begin{equation} \label{Bayesian_sender_delta}
\delta_t(t')=\left\{ \begin{array}{ll}
1 & \mbox{if } t' = t \\
0 & \mbox{otherwise}
\end{array}\right. \enspace .
\end{equation}

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's degree of rationality, i.e. to what extent the speaker sticks to the strict optimum. The cost term is used to encode preference in different utterances, be it about the utterances' lengths or syntactic categories.

@@

The pragmatic listener in the RSA model, upon receiving the utterance $m$, does a Bayes update on his prior belief $\mathcal{S}(t)$ by using his knowledge of the speaker's behavior (\ref{Bayesian_speaker}):
\begin{equation} \label{Bayesian_rec_update}
\rho(t \mid m) \propto \mathcal{S}(t)\cdot \sigma(m \mid t) \enspace .  
\end{equation}
The Bayes rule helps naturally integrate the perceptual salience of each object, which is treated as the prior $\mathcal{S}(t)$ and can be empirically measured, with listener's expectation of the speaker being informative, which is incorporated as the likelihood, thus addressing the previously mentioned challenge of balancing different sources of information. Setting $\lambda=1$ and $\mbox{Cost}(m)=0$ for all $m$, \cite{Frank} obtained a highly significant correlation between the model prediction and the experiment data.

>>>>

Referential games are confined interactive reasoning tasks used to
study pragmatic inference in a controlled environment \cite{Frank}. A
referential game consists of a set of objects called a
\textit{context}. A context contains different shapes of different
colors in various arrangements, such as shown in Figure~\ref{context}.
The speaker in the game is asked to refer to one of the objects (the
\emph{target}) by uttering an expression (typically a single word
denoting a feature, i.e., color or shape, of the target) to the
listener. The listener, who has does not know which object is the
target, needs to recover the target based on the speaker's choice of
expression.

\begin{figure}[h] 
  \centering \subfigure[Example context: green square and circle, and
  blue circle.]{\label{context}
  \begin{tikzpicture}
     \path (-1.8, 0) node [shape=rectangle, draw=green!65 ,fill= green!65 ,minimum size=28]{}
           ( 0, 0) node  [shape=circle, draw=green!65, fill= green!65,minimum size=30] {}
           ( 1.8, 0) node [shape=circle, draw=blue!65, fill=blue!65  ,minimum size=30] {}
           ( -1.8, -0.8) node {$t_{1}$}
           ( -0.0, -0.8) node {$t_{2}$}
           (  1.8, -0.8) node {$t_{3}$}
     ;

  \end{tikzpicture}
}\hspace{5mm} \subfigure[Vocabulary and truth table]{\label{VaTT}
  \begin{tikzpicture}
     \path (0,0) node[] 
     {\begin{tabular}{r|ccc}
                         & \quad $t_{1}$\ \  &\quad $t_{2}$\ \  &\quad $t_{3}$ \ \   \\
    \hline
   ``Green'':  $m_{1}$ \quad  &  $1$        &   $1$       & $0$      \\
   ``Circle'': $m_{2}$  \quad &  $0$        &   $1$       & $1$      \\ 
   ``Square'': $m_3$ \quad    &  $1$      &   $0$       & $0$      \\
   ``Blue'':   $m_4$  \quad  &  $0$        &   $0$       & $1$     
     \end{tabular}
     } ;
  \end{tikzpicture}
  }
  \caption{A simple referential game}
\end{figure}

Let us first try to play this game ourselves, to get a better sense of
what it is about and why it is relevant to human pragmatic
reasoning. For simplicity let us assume that the utterances that the
speaker is allowed to use are from a \emph{vocabulary} which is
commonly known between the speaker and the listener. The content of
the vocabulary and the (truth-conditional/literal) meaning of each
word are shown in Figure \ref{VaTT}. Suppose we play as the speaker
and the target is $t_1$, the green square. We can either use
``Square'' or ``Green'' to truthfully refer to it, but it seems more
prudent to use ``Square'', as there is only one square in the context
and thus the listener can easily identify it, whereas there are two
green objects, which makes ``Green'' ambiguous. In terms of the
Gricean \emph{Maxim of Quantity}, we should use ``Square'' because it
is more informative than ``Green'' in the given context; the surplus
informativity moreover seems relevant. Similarly, we should use
``Blue'' to refer to $t_3$, the blue circle. However, using ``Blue''
to refer to the blue circle might intuitively sound a little
unnatural, as color terms are usually used as adjectives\footnote{They
  are used as nouns mostly to refer to the colors themselves.}, while
we usually use nouns to refer to concrete objects. This inclination
becomes more evident when we want to refer to $t_2$, the green
circle. While ``Green'' and ``Circle'' are equally ambiguous in this
case, we might nevertheless prefer the latter.

Now let us turn to play as the listener. If we hear ``Square'' or
``Blue'', we easily know the intended referent as there is no
ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are
two circles in the context that we need to choose from. On the one
hand, the blue circle, having the unique color in the context, seems
to be perceptually dominant and thus easily captures our attention. On
the other hand, from the previous analysis we know that if the blue
circle were the intended referent, the speaker could have chosen
``Blue'' which is not ambiguous and thus more informative. Hence the
listener needs to balance two sources of information, i.e., the
(presumably subconscious) perceptual salience of different objects and
the rational expectation of the likelihood of the speaker making the
utterance for each object. The latter line of reasoning is crucial to
the classic Gricean account of scalar implicatures, and the major
challenges in terms of formal modeling are how to quantify notions
such as informativeness and how different pieces of information should
be integrated.

The RSA model addresses the problems by using information-theoretic
concepts to measure informativeness and Bayesian inference to
integrate different sources of information
\cite{Frank2009,Frank,Bergen2012,GoodmanStuhlmuller2013:Knowledge-and-I}. In
order to measure the informativeness of an utterance, the RSA model
starts with the literal listener who upon receiving an utterance $m$
does a conditioning on its literal meaning:
\begin{equation}\label{Bayesian_literal}
\rho_0(t \mid m) = \mathcal{U}(t\mid \intp{m}) ,
\end{equation}
where $\mathcal{U}$ is a uniform distribution over all the possible
referents. The informativeness of utterance $m$ for the intended
referent $t$ can be measured as the negative Kullback-Leibler
divergence of the induced literal listener's belief $\rho_0$ from the
speaker's own belief $\delta_t$:
\begin{equation} \label{Bayesian_sender_informativeness}
\mbox{Info}(m,t)=-\mbox{KL}(\delta_t \| \rho_0)=-\sum_{t'}\delta_t(t')\log\left(\frac{\delta_t(t')}{\rho_0(t')}\right),
\end{equation}
where $\delta_t$ is a delta distribution with all probability mass on
target object $t$, as the speaker knows her intended referent:
\begin{equation} \label{Bayesian_sender_delta}
\delta_t(t')=\left\{ \begin{array}{ll}
1 & \mbox{if } t' = t \\
0 & \mbox{otherwise}
\end{array}\right. \enspace .
\end{equation}

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum. The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

@@

The pragmatic listener in the RSA model, upon receiving the utterance
$m$, performs a Bayesian update on his prior belief $\mathcal{S}(t)$
by using an estimate of the speaker's behavior
(\ref{Bayesian_speaker}):
\begin{equation} \label{Bayesian_rec_update}
\rho(t \mid m) \propto \mathcal{S}(t)\cdot \sigma(m \mid t) \enspace .  
\end{equation}
Bayes' rule naturally integrates the perceptual salience of each
object, which is treated as the prior $\mathcal{S}(t)$ and can be
empirically measured, with listener's expectation of the speaker being
informative, which is incorporated as the likelihood, thus addressing
the previously mentioned challenge of balancing different sources of
information. Setting $\lambda=1$ and $\mbox{Cost}(m)=0$ for all $m$,
\cite{Frank} obtained a highly significant correlation between the
model prediction and their experiment data on actual speaker and
listener behavior gathered from referential games of varying
complexity.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% TeX-PDF-mode: t
%%% End:

### 2013-10-30 03_alternatives.tex
cross-ref & conclusion
rephrased intro, amended layout
Sections 2,3 proof

The RSA model has the theoretical advantage in that it is quantitative, and it has been shown to be able to fit the empirical data well. However, if we want to gain more insight about communication involving referential expressions from its success, it will be worthwhile to examine the model carefully to pin down the major components, spell out the design choices and the underlying assumptions, test their contribution to the predictive power of the model through comparison with the alternatives and scrutinize their implications thereof. In this section, we will carry out a conceptual analysis of the RSA model by comparing it to the closely related game-theoretic approaches to pragmatics, to define a class of alternative models. We will then compare the predictions of these models to the empirical data in later sections.

First and foremost, there are three closely related notions, i.e. \emph{belief}, \emph{goal} and \emph{action}, that play a crucial role in the RSA as well as game-theoretic models. The relation and distinction between them can be best illustrated by looking at (\ref{Bayesian_speaker}):
$$\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{U}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .$$
The term $\mathcal{U}(t\mid \intp{m})$ is the speaker's \emph{belief} about the listener's interpretation of the utterance. The expected utility $\mbox{U}(m,t)=\log \mathcal{U}(t\mid \intp{m})-\mbox{Cost}(m)$ describes the speaker's \emph{goal} of communication, i.e. inducing a belief that is as close as his own with minimal effort. The production rule $\sigma(m \mid t)$ on the left-hand side specifies the speaker's \emph{action}, as it defines the probability the speaker actually chooses an action, i.e. uttering a particular expression. The soft-max rule connects these notions, as a (sub-)rational agent's action depends on his belief and goal. While the RSA and game-theoretic models all share this general architecture in their designs, they vary in the specific interpretations of the three ingredients, which reflect different views and emphasis on language and communication and give rise to a series of interesting and important empirical questions. Again, let us continue with the speaker model (\ref{Bayesian_speaker}) to illustrate some points of divergence and formally define the corresponding alternatives.

First of all, although \cite{Frank} claims that the empirically measured perceptual salience is used to capture the notion of common knowledge between the speaker and the listener, it turns out that the speaker's production rule (\ref{Bayesian_speaker}) simply does not take the listener's salience prior $\mathcal{S}(t)$ into account in her literal listener model (\ref{Bayesian_literal}). This makes it unclear in what sense the speaker knows the listener's perceptual salience. A natural variant, which is used in some game-theoretic models, would be to replace the literal listener's uniform distribution $\mathcal{U}(t)$ in (\ref{Bayesian_literal}) with the salience prior $\mathcal{S}(t)$. This will lead to the alternative production rule:
\begin{equation} \label{Bayesian_speaker_salience}
\sigma_\mathcal{S}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{S}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .
\end{equation}

Secondly, the informativeness of an utterance $m$, which is a crucial part of the communication goal, is measured in the RSA model in terms of how close the induced belief of the literal listener is from that of the speaker's own belief. This means that the RSA model sees the goal of communication as conveying belief. While it is normally true that language does convey the belief of the speaker, it is questionable at least in this referential scenario whether letting the listener form a proper belief is the ultimate goal of the communication. After all, if the speaker wants to refer to something, it seems that in the end what matters is whether the listener actually picks out the intended referent successfully (e.g., when the speaker wants the listener to pass something). We might call this view \emph{action-oriented}, in contrast to the \emph{belief-oriented} view of communication which the RSA model adopts, as it interprets the goal of communication as invoking the intended action rather than forming an accurate belief. \footnote{Note that the two views are definitely intimately related, despite our emphasis on their differences here. An accurate belief is usually helpful to choosing a desired action. In fact, forming a belief can itself be seen as an action that the listener performs, which might well be all that the speaker cares about in many everyday situations, e.g. casual talks or responses to inquiries of information. Also, the relevant actions might vary greatly across situations which makes belief formation a practical alternative to serve as the goal. Nevertheless, they are still essentially different views in spite of the fact that they often coincide in practice. The referential communication scenario we investigate in this paper reflects the distinction.} Thus, according to this view, the informativeness of a message would be measured as the probability of the listener choosing the intended referent. Formally speaking, we have the action-oriented speaker model

@@

Hence we have four types of speaker models that differ in either the speaker's belief about the literal listener, or the speaker's goal of communication. We now introduce a uniform notation $\sigma_{xy},\ x\in\{a,b\},y\in\{\mathcal{U},\mathcal{S}\}$ for them:
\begin{equation} \label{Bayesian_speaker_uniform_action}
\sigma_{\mathrm{a}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot ( y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
\begin{equation} \label{Bayesian_speaker_uniform_belief}
\sigma_{\mathrm{b}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
where $\mathcal{U}$ is the uniform prior and $\mathcal{S}$ is the salience prior. For example, in the original RSA model, the speaker does not take listener's salience prior into account and he has a belief-oriented goal of communication, thus it will be denoted as $\sigma_{\mathrm{b}\mathcal{U}}$

Finally, the original RSA speaker model in \cite{Frank} has $\mbox{Cost}(m)=0$ for all $m$, which means that the potential speaker preference in different utterances is not taken into account. As we previously pointed out, our intuition seems to suggest a preference for nouns over adjectives as referential expressions. Since it is an empirical question whether, to what direction or to what extent such a preference exists, we leave open all the possibilities. Technically speaking, since only the difference in costs matters, we define the cost function using a constant $c\in \mathbb{R}$
\begin{equation} \label{cost}
\mbox{Cost}(m)=c \mbox{ if } m \mbox{ is an adjective and }0 \mbox{ otherwise} \enspace . 
\end{equation}
If $c>0$ it means there is a preference for nouns and if $c<0$ then the preference is for adjectives. No preference exists if $c=0$.

@@

Let us start from the relatively clear part. As pointed out previously, the likelihood term $\sigma(m \mid t)$ is the listener's belief about how the speaker behaves. One natural question is to what extent the listener's belief correlate with the actual production. We thus treat the speaker's production term $\sigma(m \mid t)$ in the listener's model as a parameter, making the association with a belief about production of the listener model explicit:
\begin{equation} \label{listener-production}
\rho(\sigma_{xy})(t \mid m) \propto \mathcal{S}(t)\cdot \sigma_{xy}(m \mid t) \enspace .
\end{equation}
Note that the speaker's production rule has two parameters $\lambda_\mathrm{S},c$ which are also be included in the above specification of the listener model.

@@

Finally, since the RSA model adopts a belief-oriented view on the goal of communication, its listener model only consists of the belief about the intended referent after hearing the utterance, obtained by a Bayes update. However, according to the action-oriented view, this is not quite the end of the story. The listener often needs to decide the exact referent of a referential expression. (Again, consider the case in which the listener is asked to pass something.) Hence in such cases the listener will choose an object that he (soft-)maximally believes to be the intended referent. Formally speaking, the action-oriented listener model would be
\begin{equation} \label{listener-action}
\rho_{\mathrm{a}v}(\sigma_{xy})(t \mid m) \propto \exp(\lambda_\mathrm{L} \cdot \rho_{\mathrm{b}v}(\sigma_{xy})(t \mid m) ),
\end{equation}
where $v\in\{\mathcal{U},\mathcal{S}\}$, $\lambda_\mathrm{L}$ is the parameter measuring the listener's degree of rationality, and $\rho_{\mathrm{b}v}$ is the belief-oriented model that does the Bayes update:
\begin{equation} \label{listener-belief}
\rho_{\mathrm{b}v}(\sigma_{xy})(t \mid m) \propto v(t)\cdot \sigma_{xy}(m \mid t) \enspace .
\end{equation}
For instance, the original RSA listener model is a belief-oriented one with the perceptual salience as prior, whose belief about the speaker is $\sigma_{\mathrm{b}\mathcal{U}}$, hence it is denoted as $\rho_{\mathrm{b}\mathcal{S}}(\sigma_{\mathrm{b}\mathcal{U}})$.

>>>>

The RSA model has a theoretical advantage over traditional formal
pragmatic theories in that it provides quantitative predictions, and
it has been shown to fit relevant empirical data well. However, if we
want to learn more about pragmatic reasoning about referential
expressions, it will be worthwhile to examine RSA carefully to pin
down its major components, to spell out its main design choices and
their underlying assumptions, and to test their contribution to the
predictive power of the model through statistical comparison with
natural alternatives. In this section, we will therefore compare the
RSA model to closely related game theoretic approaches, that likewise
assume that speakers and listeners form cascading beliefs about mutual behavior
and seek to optimize their behavior based on these beliefs
\cite{Rabin1990:Communication-b,Stalnaker:SayingMeaningCredibility,BenzvanRooijOptimalAssertions2007,Franke2011:Quantity-Implic,Jager2013:Rationalizable-}.\footnote{See
\cite{FrankeJager2013:Pragmatic-Back-} for overview and comparison
of game theoretic and Bayesian models.} Doing so
let's us describe a class of potential alternatives to RSA. We will
then compare the predictive power of these given relevant empirical
data in later sections.

There are three closely related notions, i.e., \emph{belief},
\emph{goal} and \emph{action}, that play a crucial role in RSA as well
as game theoretic models. The relation and distinction between them
can be best illustrated by looking at (\ref{Bayesian_speaker}),
repeated here:
$$\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{U}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .$$
The term $\mathcal{U}(t\mid \intp{m})$ is the speaker's \emph{belief}
about the listener's interpretation of the utterance. The expected
utility $\mbox{U}(m,t)=\log \mathcal{U}(t\mid
\intp{m})-\mbox{Cost}(m)$ describes the speaker's \emph{goal} of
communication, i.e., inducing a belief that is as close as possible to
her own with minimal effort. The production rule $\sigma(m \mid t)$ on
the left-hand side specifies the speaker's \emph{action}, as it
defines the probability with which, according to RSA, speakers would
choose a particular expression. The soft-max rule connects these
notions, as a (sub-)rational agent's action depends on his belief and
goal. While the RSA and game theoretic models all share this general
architecture in their designs, they vary in the specific
interpretations of the three ingredients, which reflect different
views and emphasis on language and communication and give rise to a
series of interesting and important empirical questions. Again, let us
continue with the speaker model (\ref{Bayesian_speaker}) to illustrate
some points of divergence and formally define the corresponding
alternatives.

First of all, although it is prima facie reasonable to hypothesize, as
\cite{Frank} do, that the empirically measured perceptual salience
$\mathcal{S}(\cdot)$ is common knowledge between speaker and listener,
it does not actually affect RSA's production rule
(\ref{Bayesian_speaker}) at all. This makes it unclear in what sense
the speaker can be said to know the listener's perceptual salience,
since neither his beliefs nor actions depend on it. A natural variant,
which is used in some game theoretic models, would replace the literal
listener's uniform distribution $\mathcal{U}(t)$ in
(\ref{Bayesian_literal}) with the salience prior
$\mathcal{S}(t)$. This leads to the alternative production rule:
\begin{equation} \label{Bayesian_speaker_salience}
\sigma_\mathcal{S}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{S}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .
\end{equation}

Secondly, RSA measures the informativeness of an utterance $m$, which
is a crucial part of the communicative goal, in terms of how close the
induced belief of the literal listener is from the speaker's own
belief. This means that the RSA model sees the goal of communication
as conveying belief. While it is normally true that language does
convey the belief of the speaker, it is questionable at least in this
referential scenario whether letting the listener form a proper belief
is the ultimate goal of the communication. After all, if the speaker
wants to refer to something, it seems that in the end what matters is
whether the listener actually picks out the intended referent
successfully (e.g., when the speaker wants the listener to pass
something). This view is inherent in game theoretic approaches where
agents' beliefs are backed up by explicit formulations of their
utilities. We might call this latter view \emph{action-oriented}, in
contrast to the \emph{belief-oriented} view of communication which the
RSA model adopts, as it interprets the goal of communication as
invoking the intended action rather than forming an accurate
belief.\footnote{Note that the two views are clearly intimately
  related, despite our emphasis on their differences here. An accurate
  belief is usually helpful for choosing a desired action. In fact,
  forming a belief can itself be seen as an action that the listener
  performs, which might well be all that the speaker cares about in
  many everyday situations, e.g., casual talks or responses to
  inquiries of information. Also, the relevant actions might vary
  greatly across situations which makes belief formation a practical
  alternative to serve as the goal. Nevertheless, they are still
  essentially different views in spite of the fact that they often
  coincide in practice. The referential communication scenario we
  investigate in this paper reflects the distinction.} Thus, according
to this view, the informativeness of a message would be measured as
the probability of the listener choosing the intended
referent. Formally speaking, we have the action-oriented speaker model

@@

Hence we have four types of speaker models that differ in either the speaker's belief about the literal listener, or the speaker's goal of communication. We now introduce a uniform notation $\sigma_{xy},\ x\in\{a,b\},y\in\{\mathcal{U},\mathcal{S}\}$ for them:
\begin{equation} \label{Bayesian_speaker_uniform_action}
\sigma_{\mathrm{a}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot ( y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
\begin{equation} \label{Bayesian_speaker_uniform_belief}
\sigma_{\mathrm{b}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
where $\mathcal{U}$ is the uniform prior and $\mathcal{S}$ is the
salience prior. For example, in the original RSA model, the speaker
does not take listener's salience prior into account and he has a
belief-oriented goal of communication. Thus it will be denoted as
$\sigma_{\mathrm{b}\mathcal{U}}$.

Finally, the original RSA speaker model of \cite{Frank} has
$\mbox{Cost}(m)=0$ for all $m$, which means that the potential speaker
preference in different utterances is not taken into account (but not
so in, e.g., \cite{Bergen2012}). As we previously pointed out, our
intuition seems to suggest a preference for nouns over adjectives as
referential expressions.\footnote{The rich tradition on referential
  expression generation \cite{KramerDeemter2012:Computational-G} has
  also identified the need to take speaker preferences into account,
  and \cite{GattGompel2013:Are-we-Bayesian} criticize RSA on these
  grounds as well.} Since it is an empirical question whether, to what
direction or to what extent such a preference exists, we leave open
all the possibilities. Technically speaking, since only the difference
in costs matters, we define the cost function using a constant $c\in
\mathbb{R}$
\begin{equation} \label{cost}
\mbox{Cost}(m)=c \mbox{ if } m \mbox{ is an adjective and }0 \mbox{ otherwise} \enspace . 
\end{equation}
If $c>0$ it means there is a preference for nouns and if $c<0$ then the preference is for adjectives. No preference exists if $c=0$.

@@

Let us start from the relatively clear part. As pointed out
previously, the likelihood term $\sigma(m \mid t)$ is the listener's
belief about how the speaker behaves. One natural question is to what
extent the listener's belief correlates with the actual production. We
thus treat the speaker's production term $\sigma(m \mid t)$ in the
listener's model as a parameter, making the association with a belief
about production of the listener model explicit:
\begin{equation} \label{listener-production}
\rho(\sigma_{xy})(t \mid m) \propto \mathcal{S}(t)\cdot \sigma_{xy}(m \mid t) \enspace .
\end{equation}
Note that the speaker's production rule has two parameters
$\lambda_\mathrm{S},c$ which are also included in the above
specification of the listener model.

@@

Finally, since the RSA model adopts a belief-oriented view on the goal
of communication, its listener model only consists of the belief about
the intended referent after hearing the utterance, obtained by a Bayes
update. However, according to the action-oriented view, this is not
quite the end of the story. The listener often needs to decide the
exact referent of a referential expression. (Again, consider the case
in which the listener is asked to pass something.) Hence in such cases
the listener will choose an object by, essentially, (soft-)maximizing
over his beliefs. Formally, the action-oriented listener model
becomes:
\begin{equation} \label{listener-action}
\rho_{\mathrm{a}v}(\sigma_{xy})(t \mid m) \propto \exp(\lambda_\mathrm{L} \cdot \rho_{\mathrm{b}v}(\sigma_{xy})(t \mid m) ),
\end{equation}
where $v\in\{\mathcal{U},\mathcal{S}\}$, $\lambda_\mathrm{L}$ is the
parameter measuring the listener's degree of rationality, and
$\rho_{\mathrm{b}v}$ is the belief-oriented model that does the
Bayesian update:
\begin{equation} \label{listener-belief}
\rho_{\mathrm{b}v}(\sigma_{xy})(t \mid m) \propto v(t)\cdot \sigma_{xy}(m \mid t) \enspace .
\end{equation}
For instance, the original RSA listener model is a belief-oriented one with the perceptual salience as prior, whose belief about the speaker is $\sigma_{\mathrm{b}\mathcal{U}}$, hence it is denoted as $\rho_{\mathrm{b}\mathcal{S}}(\sigma_{\mathrm{b}\mathcal{U}})$.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% TeX-PDF-mode: t
%%% End:
